The title

### 2024-09-01 05_Manuscript/manuscript.Rmd
work on intro

Introduction

Example citation: [@dplyr]

[@rosseel2012]

Intro - awareness of death, especially in light of COVID.

Various existential philosophers and psychologists have proposed ways in which we deal with the awareness of death and the anxiety this awareness often causes. Psychoanalyst Erik Erikson (1950) proposed that during mid-life one becomes acutely aware of their oncoming death and is motivated to care for things which will outlast themselves. He called this act of caring generativity. In *The Denial of Death* (1973), philosopher Ernest Becker posits that humans undertake immortality projects to curb their sense of vulnerability to death. Similarly, psychiatrist Robert Jay Lifton (1979), a mentee of Erikson, described the awareness of death as being ever present and motivating us to create symbols thereby allowing ourselves to imagine us as symbolically immortalized. Existential psychiatrist, Irvin Yalom (2008) notes that many of his clients experiencing anxiety about their death take comfort in “rippling”, the idea that one’s lasting effects on the world will ripple out and influence the world after they have died.

Although these thinkers use different terminology, there are several common themes among their ideas. (1) Our physical death is an inevitability, and we often find our awareness of its inevitability to be aversive: that is, it can cause angst, death-anxiety, despair, being-towards-death, terror, and so on. However, (2) we take comfort in the idea that other, non-physical parts of us continue to exist indefinitely after our biological death, through mechanisms such as the heroic archetype and symbolic self. (3) Finally, we can take action to promote these non-physical aspects of the self, such as through search for meaning, immortality project, care, generativity, rippling.

Sense of symbolic immortality was originally theorized by Lifton (1979), who thought that awareness of death drives a fundamental human need to "develop a sense of continuity and lastingness" beyond the lifespan (quote from Drolet 1990). Humans as meaning-seeking creatures.

@@

Sense of psychological death?

Though some denial of death is thought to be universal, acceptance of the reality of death can drive what Drolet (1990) refers to as personal and collective creativity, or what Erikson referred to as generativity. Erikson's theory and generativity.

Current measures of generativity are more about domains of generativity, actions, trait rather than underlying motives. Loyola Generativity Scale and the authors' understanding of the dimensions of generativity. Underlying motivation of generativity thought to be related to a motive to expand the sense of self beyond the lifetime (Kotre 1984), especially in light of the fear of death. McAdams and de St. Aubin (1992) indeed found symbolic immortality to be one of five themes across content analysis of autobiographical episodes.

Somewhat loss of religiosity in the West that may have been a way of coping/finding a sense of immortality in the past. Likely this may have fallen under the umbrella of the spiritual mode of immortality, and for many a literal belief in life after death. Perhaps many now would benefit more from a sense of collective, social, or creative immortality. (some citations here if I keep this paragraph)

Claims of how we suppress death-anxiety have been investigated experimentally, primarily through the paradigm of Terror Management Theory (TMT). Based on the theories of Ernest Becker, TMT posits that human awareness of death is always present to some degree. This awareness of our inevitable death, coupled with a strong aversion to thoughts of death, causes terror and is pushed out of our consciousness by our creation of meaning systems (CITATION). TMT refers to a person's awareness of death as mortality salience (MS). The MS hypothesis of TMT posits than an increase in one's awareness of death causes an increase in compensatory behaviors to lower their death anxiety, either by distracting from the awareness of death or by the promotion of meaningful cultural worldviews. A meta-analysis of 277 experiments has found mortality salience to have moderate effects (*r* = 0.35; Burke, Martens & Faucher, 2010). TMT literature has found three consistent buffers of our awareness of death: self-esteem, interpersonal relationships, and cultural worldview (CITATION, unless it's the same as above, in which case make that clear).

TMT research is conducted in two paradigms. In the MS paradigm, a participant’s increased awareness of death causes an increase in compensatory buffers. In the death-thought accessibility paradigm, experimentally decreasing participants’ buffers, such as lowering self-esteem or challenging their worldview beliefs, causes (CHECK THIS AND CITE ACTUAL STUDIES) an increase in their awareness of death. Together, these experiments provide convincing evidence of these of suppressing awareness of death and death anxiety.

It is assumed that such changes in attitude and behavior suppress thoughts of death by giving a sense of existential immortality; however, to our knowledge no researchers have attempted to assess subjective feelings of immortality. This missing literature may in part be because of a lack of a viable measure of existential immortality. We view this dearth as a large gap in the literature and methodology. Two attempts have been made to develop such a measurement: Drolet's (1990) Sense of Symbolic Immortality Scale and Mathews and Kling (1988) measure of symbolic immortality, based on an original questionnaire by Mathews and Mister (1987).

NOTE\*\*\* can't get access to the Mathews and Kling article but it seems mostly related to prosocial behavior for nonprofit stuff - should we just focus on the original Mathews and Mister article? Some other articles I found use the Mathews and Kling version though. <https://journals.sagepub.com/doi/abs/10.1177/089976408801700202?download=true&journalCode=nvsa>

@@

Mathews and Mister (1987) also developed a scale involving symbolic immortality, sensation seeking, and psychic numbness. Some studies have used a revised version of the scale by Mathews and Kling (1988), who adapted it for a study on prosocial behavior in the context of nonprofit volunteer motivation. Mathews and Mister's study included 400 adults. Experiential transcendence was operationalized as something like Zuckerman's (1979) sensation seeking, which may not fully capture the original intent (the experience of losing oneself). Items were forced into five factors, and the five factors largely aligned with Lifton's constructs. Although internal consistency was at least acceptable for each factor, goodness of fit statistics are not reported. The items themselves seem to have more face validity than the scale by Drolet, but some factors seem more behavioral and specific: pertaining to one's religiosity or biological children, whereas Lifton's theory allows for a broader interpretation of these dimensions. The Nature and Creative factors seems most useful and theoretically aligned with Lifton.

The goal of the present research is to develop a scale that more directly measures one’s sense of existential immortality in a way more generally applicable regardless of a person's religious beliefs and parental status. We have attempted to use current best practices for scale development.

>>>>

Intro paragraph - awareness of death, especially in light of COVID.

Various existential philosophers and psychologists have proposed ways in which we deal with the awareness of death and the anxiety this awareness often causes. Psychoanalyst Erik Erikson (1950) proposed that during mid-life one becomes acutely aware of their oncoming death and is motivated to care for things which will outlast themselves. He called this act of caring generativity. In *The Denial of Death*, philosopher Ernest Becker (1973) posits that humans undertake immortality projects to curb their sense of vulnerability to death. Similarly, psychiatrist Robert Jay Lifton (1979), a mentee of Erikson, described the awareness of death as being ever present and motivating us to create symbols, thereby allowing us to imagine ourselves as symbolically immortalized. Existential psychiatrist Irvin Yalom (2008) notes that many of his clients experiencing anxiety about their death take comfort in “rippling,” the idea that one’s lasting effects on the world will ripple out and influence the world after they have died.

Although these thinkers use different terminology, there are several common themes among their ideas. (1) Our physical death is an inevitability, and we often find our awareness of its inevitability to be aversive. This aversion may be referred to varyingly as angst, death-anxiety, despair, being-towards-death, terror, and so on. However, (2) we take comfort in the idea that other, non-physical parts of us continue to exist indefinitely after our biological death, through mechanisms such as the heroic archetype and symbolic self. (3) Finally, we can take action to promote these non-physical aspects of the self, such as through search for meaning, immortality project, care, generativity, and rippling.

One of these bodies of thought, called symbolic immortality, was originally theorized by Lifton (1979), who thought that awareness of death drives a fundamental human desire for a sense of continuity lasting beyond the lifespan. Essentially, humans are meaning-seeking creatures, and throughout our lives, this search for meaning involves an evolving psychological imagery of life and death. Death, or the transient nature of life, threatens our search of meaning. He thought that if we could achieve what we believe to be some form of immortality, we could overcome this loss of meaning, and the awareness of death could instead drive an inner vitality (imagery associated with connection, integrity, and movement). If this drive toward vitality is lost, we are vulnerable to what Lifton called a psychic numbness (imagery associated with separation, disintegration, and stasis). In Lifton's own words, "Death does indeed bring about biological and psychic annihilation. But life includes symbolic perceptions of connections that precede and outlast that annihilation" (1979, p. 18).

@@

Claims of how we suppress death-anxiety have been investigated experimentally, primarily through the paradigm of Terror Management Theory (TMT). Based on the theories of Ernest Becker, TMT posits that human awareness of death is always present to some degree. This awareness of our inevitable death, coupled with a strong aversion to thoughts of death, causes terror and is pushed out of our consciousness by our creation of meaning systems [@greenberg1986]. TMT proposes that self-esteem, interpersonal relationships, and cultural worldview work together to buffer against our anxiety about death. It is assumed that these buffers suppress thoughts of death by providing a sense of symbolic immortality, though little systematic research has been conducted on this construct. The results of this buffering process are not always positive. For example, experimentally priming mortality salience can lead to more positive attitudes toward in-group members but harsher negative attitudes toward out-group members [@greenberg1990].

TMT refers to a person's awareness of death as mortality salience (MS). The MS hypothesis of TMT posits than an increase in one's awareness of death causes an increase in compensatory behaviors to lower their death anxiety, either by distracting from the awareness of death or by the promotion of meaningful cultural worldviews. In the MS paradigm, experimentally priming a participant’s awareness of death (for example, by having participants write about death and then complete a distraction task) is thought to cause an increase in compensatory buffers. A meta-analysis of 277 experiments found mortality salience to have a robust, moderate overal effect size: *r*(276) = 0.35, *p* = .00 [@burke2010]. Altogether, these experiments provide convincing evidence for TMT and the MS hypothesis in particular.

Though some avoidance of (or buffering against) death anxiety is thought to be universal and has the potential to increase interpersonal conflict, awareness of death through symbolic immortality may also have potential as a positive force. In particular, it is thought to be an underlying motive for what Erikson referred to as generativity. Generativity is the seventh of eight proposed stages in Erikson's (1950) theory of psychosocial development, which he associated with midlife and described as "the concern in establishing and guiding the next generation" (Erikson, 1963, p. 267). Little systematic research was conducted on this subject until the 1980's. Kotre (1984) expanded on the theory and proposed that the drive for generativity was related to a motive to expand the sense of self beyond the lifetime, especially in light of the fear of death.

McAdams and de St. Aubin (1992) sought to formalize the study of generativity as a multidimensional construct. Their seven components of generativity include cultural demand, inner desire (for symbolic immortality and community), concern (for the next generation), belief (in the human species), commitment, action, and narration (of generativity within one's life story). In addition to a quantitative measure of generative concern (the Loyola Generativity Scale), they developed a system for content analysis of autobiographical episodes pertaining to generativity, and symbolic immortality is one of the five themes they found. Here they define symbolic immortality as "any reference to leaving a legacy, having an enduring influence, or leaving behind products that will outlive one's physical existence," a theme clearly related to both Lifton's and Erikson's theories (McAdams & de St. Aubin, 1992, p. 1011).

These research areas depend on the construct of symbolic immortality for their theoretical frameworks, but few researchers have attempted to systematically and quantitatively assess this construct. Two attempts have been made to develop such a measurement: Drolet's (1990) Sense of Symbolic Immortality Scale and Mathews and Kling (1988) measure of symbolic immortality, based on an original questionnaire by Mathews and Mister (1987).

@@

Mathews and Mister (1987) also developed a scale involving symbolic immortality, sensation seeking, and psychic numbness. Some studies have used a revised version of the scale by Mathews and Kling (1988), who adapted it for a study on prosocial behavior in the context of nonprofit volunteer motivation. Mathews and Mister's study included 400 adults. Experiential transcendence was operationalized as something like Zuckerman's (1979) sensation seeking, which may not fully capture the original intent (the experience of losing oneself). Items were forced into five factors, and the five factors largely aligned with Lifton's constructs. Although internal consistency was at least acceptable for each factor, goodness of fit statistics are not reported. Mathews and Kling's (1988) study reported similar results for their revised scale. The items themselves seem to have more face validity than the scale by Drolet, but some factors seem more behavioral and unnecessarily specific: pertaining to one's religiosity or biological children, whereas Lifton's theory allows for a broader interpretation of these dimensions. The Nature and Creative factors seem most useful and theoretically aligned with Lifton.

Much more advanced factor analysis methods have been developed since the 1980s, but to our knowledge, these scales have not been tested with more robust tools. The goal of the present research is to develop an up-to-date symbolic immortality scale that more directly measures one’s sense of symbolic immortality and which contains items more generally applicable to broad groups of participants (e.g., regardless of a person's religious beliefs and parental status). We have attempted to use current best practices for scale development and analysis.
